Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-27 ind_files/03_BD_processes.tex
Move examples to appendix

\subsection{Density-dependence and the intuition for system-size expansions in ecology}
The fundamental idea behind the system-size expansion relates to the nature of the jumps between successive states of a birth-death process. In most situations in ecology, at an individual level, births and deaths of individuals are affected by local population density and not directly by the total population size. Despite this, the jumps themselves occur in terms of the addition (birth) or removal (death) of a \emph{single individual} from the population. If there are many individuals, each individual contributes a negligible amount to the density, and thus, the discontinuous jumps due to individual-level births or deaths can look like a small, \emph{continuous} change in population density. This is the essential idea behind the system-size expansion. The name derives from the formalization of this idea as a change of variable from the discrete values $\{0,1,2,\ldots,n-1,n,n+1,\ldots\}$ to the approximately continuous values $\{0,1/K,2/K,\dots,x-1/K,x,x+1/K,\ldots\}$ by the introduction of a `system size parameter' $K$. In ecology, this parameter will be some fundamental limit on resources, such as habitat size or carrying capacity. In physics and chemistry, it is usually the total volume of a container in which physical or chemical reactions take place. When $K$ is large, the fact that transitions occur in units of a small value $1/K$ can be exploited via a Taylor expansion of the transition rates in the Master equation, which then yields a Fokker-Planck equation upon neglecting higher order terms. A similar approximation is well-known (ever since Fisher) in theoretical population genetics \citep{ewens_mathematical_2004}, where it goes by the name of the `diffusion approximation', and has been heavily used by Kimura \citep{crow_introduction_1970} in his stochastic models. However, the population genetics version of the approximation often lacks an explicit system size parameter (in physics parlance \citep{gardiner_stochastic_2009}, it is closer to a Kramers-Moyal expansion than a Van Kampen expansion) and is thus often somewhat ad-hoc.

>>>>

\subsection{Density-dependence and the intuition for system-size expansions in ecology}
The fundamental idea behind the system-size expansion relates to the nature of the jumps between successive states of a birth-death process. In most situations in ecology, at an individual level, births and deaths of individuals are affected by local population density and not directly by the total population size. Despite this, the jumps themselves occur in terms of the addition (birth) or removal (death) of a \emph{single individual} from the population. If there are many individuals, each individual contributes a negligible amount to the density, and thus, the discontinuous jumps due to individual-level births or deaths can look like a small, \emph{continuous} change in population density. This is the essential idea behind the system-size expansion. The name derives from the formalization of this idea as a change of variable from the discrete values $\{0,1,2,\ldots,n-1,n,n+1,\ldots\}$ to the approximately continuous values $\{0,1/K,2/K,\dots,x-1/K,x,x+1/K,\ldots\}$ by the introduction of a `system size parameter' $K$. In ecology, this parameter will be some fundamental limit on resources, such as habitat size or carrying capacity. In physics and chemistry, it is usually the total volume of a container in which physical or chemical reactions take place. When $K$ is large, the fact that transitions occur in units of a small value $1/K$ can be exploited via a Taylor expansion of the transition rates in the Master equation, which then yields a Fokker-Planck equation upon neglecting higher order terms. A similar approximation is well-known (ever since Fisher) in theoretical population genetics \citep{ewens_mathematical_2004}, where it goes by the name of the `diffusion approximation', and has been heavily used by Kimura \citep{crow_introduction_1970} in his stochastic models. However, the population genetics version of the approximation usually either relies on total population size being fixed \citep{crow_introduction_1970, lande_natural_1976,ewens_mathematical_2004} or is conducted in an ad-hoc manner without specifying an explicit system size parameter (\emph{i.e.} is closer to a Kramers-Moyal expansion than a Van Kampen expansion).

### 2023-02-27 ind_files/99_Appendix_D_examples.tex
Move examples to appendix

\epigraph{\justifying Our future advances will not be concerned with universal laws, but instead with universal approaches to tackling particular problems}{Peter Kareiva}

In the previous chapter, we used various tools and arguments to illustrate how to obtain approximate equations for very general stochastic birth-death processes. In this chapter, we put this theory into use through examples to study extinction and diversity in finite populations.

\section{An example in one dimension: The stochastic logistic equation}
Consider the functional forms of example \ref{ex_1D_stoch_logistic}, given by
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dB_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we
obtain the logistic equation. Letting $\alpha(t)$ be the solution of the logistic equation \eqref{ex_1D_stoch_logistic_det_limit}, We can Taylor expand $A^{\pm}(x)$ for the weak noise approximation, and we find:
\begin{align*}
A^-_1(x) &= \frac{d}{dx}(rx(1-x))\biggl{|}_{x=\alpha} = r(1 - 2\alpha(t))\\
A^+_0(x) &= \alpha(t)(v+r\alpha(t))
\end{align*}
Thus, the weak noise approximation of \ref{ex_1D_stoch_logistic_BD_eqns} is given by
\begin{equation}
    X_t = \alpha(t) + \frac{1}{\sqrt{K}}Y_t
\end{equation}
where the stochastic process $Y_t$ is an Ornstein-Uhlenbeck process given by the solution to the linear SDE
\begin{align}
    dY_t &= A^-_1(t)Y_tdt + \sqrt{A^+_0(t)}dB_t\nonumber\\
    \Rightarrow dY_t &= r(1 - 2\alpha(t))Y_tdt + \sqrt{\alpha(t)(v+r\alpha(t))}dB_t\label{ex_1D_stoch_logistic_WNA}
\end{align}

@@

Let us imagine a population with $m$ types of individuals that are competing for resources. Let the state of the population be characterized by the vector $\mathbf{v}(t) = [v_1(t),v_2(t),\ldots,v_m(t)]$, where $v_i(t)$ is the number of type $i$ individuals at time $t$. Let the birth and death rates of the $i$th type be given by:

@@

b_i(\mathbf{v}) &= \lambda v_i\\
d_i(\mathbf{v}) &= \left(\mu + \frac{1}{K}\left(\sum\limits_{j=1}^{m}M_{ij}v_j\right)\right)v_i

@@

where $K > 0$ is our system size parameter (and represents a global carrying capacity across all types), $\lambda > 0$ and $\mu > 0$ are suitable positive constants representing the baseline natality and mortality common to all types, and $M_{ij} \geq 0$ is a non-negative constant describing the effect of type $j$ individuals on the death rate of type $i$ individuals due to competition. We assume that $M_{ij} \ll K$. The values $M_{ij}$ are often collected in an $m \times m$ matrix $\mathbf{M}$. The negative of this matrix, $-\mathbf{M}$, is called the `payoff matrix' (in evolutionary game theory) or `interaction matrix' (in Lotka-Volterra models), which is why the infinite population analogues of such models are called `matrix games' in the evolutionary game theory literature. Lotka-Volterra models also frequently assume that the diagonal elements $M_{ii}$ are all equal, though we will not make that assumption here.\\

@@

b^{(K)}_i(\mathbf{x}) &= \lambda x_i\\
d^{(K)}_i(\mathbf{x}) &= \left(\mu + \sum\limits_{j=1}^{m}M_{ij}x_j\right)x_i

@@

A^{\pm}_{i} = x_i\left(\lambda \pm \left(\mu + \sum\limits_{j=1}^{m}M_{ij}x_j\right)\right)

@@

Defining $r = \lambda - \mu$ and $v = \lambda + \mu$, we see from equation \eqref{nD_Ito_SDE} that the mesoscopic view is the $m$ dimensional SDE given by

@@

\mathbf{A^-}_i = {(\mathbf{X}_{t})}_i(r - \sum\limits_{j=1}^{m}M_{ij}{(\mathbf{X}_{t})}_j) 

@@

(\mathbf{D}\mathbf{D}^{\mathrm{T}})_i = {(\mathbf{X}_{t})}_i(v + \sum\limits_{j=1}^{m}M_{ij}{(\mathbf{X}_{t})}_j) 

@@

\frac{d x_i}{dt} = x_i\left(r - \sum\limits_{j=1}^{m}M_{ij}x_j\right)

@@

These are precisely the competitive Lotka-Volterra equations for a system of $m$ species. By matching the terms of \eqref{nD_example_density_b_d_rates} with those of \eqref{nD_functional_forms_for_replicator}, we can identify that we have $\mu = 0$ and

@@

b^{(\textrm{int})}_{i}(\mathbf{x}) &= \lambda\\
d^{(\textrm{int})}_i(\mathbf{x}) &= \mu + \sum\limits_{j=1}^{m}M_{ij}x_j\\
w_i(\mathbf{x}) &= r - \sum\limits_{j=1}^{m}M_{ij}x_j

@@

&= \sum\limits_{i=1}^{m}\left(r - \sum\limits_{j=1}^{m}M_{ij}x_j\right)p_i\\
&= r - \sum\limits_{i=1}^{m}p_i\left(\sum\limits_{j=1}^{m}M_{ij}x_j\right)

@@

\frac{dp_i}{dt} = \left[\sum\limits_{i=1}^{m}p_i\left(\sum\limits_{j=1}^{m}M_{ij}x_j\right) - \sum\limits_{j=1}^{m}M_{ij}x_j\right]p_i

@@

Let us define the payoff matrix $\mathbf{B} = - \mathbf{M}$. Then, the dynamics are
\frac{1}{N_K(t)}\frac{dp_i}{dt} = \left[(\mathbf{Bp})_i - \mathbf{p}\cdot\mathbf{Bp}\right]p_i   
which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $B$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks. We briefly note here that if $m$ is large, then both the stochastic dynamics \eqref{nD_example_SDE} and the deterministic limit \eqref{nD_example_det_limit} can be simplified from an $m$ dimensional system to an $m-1$ dimensional system by a coordinate transformation. If we go from the variables $x_1,\ldots,x_m$ to the variables $p_1,\ldots,p_{m-1},N_K$, then, we can exploit the fact that $N_K$ varies much less than the $p_i$ terms to project the system onto a `slow manifold' in which $N_K$ is approximately constant, thus obtaining an $m-1$ dimensional system of equations and recovering the relation between the Lotka-Volterra equations for $m$ species and the replicator equation for $m-1$ tactics \citep{constable_mapping_2017,parsons_dimension_2017}. However, we will not explore this further in this manuscript, and refer the reader to \citep{constable_stochastic_2013} and \citep{parsons_dimension_2017} for a review of the ideas of (stochastic) dynamics on slow manifolds.\\

@@

&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_k\left(\frac{\partial}{\partial x_k}(r+\sum\limits_{j=1}^{m}B_{ij}x_j)\bigg{|}_{\mathbf{x}=\mathbf{a}(t)}\right)\\
&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_kB_{ik}\\

@@

where we have used the fact that $w_i(\mathbf{y}) = r + \sum\limits_{k=1}^{m}y_kB_{ik}$ (from \eqref{nD_example_fitness}) in the last step. Thus, in the weak noise approximation of our process, the dynamics are given by

@@

\resizebox{1.1\textwidth}{!}{$\displaystyle\frac{\partial P}{\partial t}(\mathbf{y},t) = \sum\limits_{i=1}^{m}\left(-\frac{\partial}{\partial y_i}\left\{\left(y_iw_i(\mathbf{a}) + a_iw_i(\mathbf{y}) - ra_i\right)P(\mathbf{y},t)\right\}+\frac{1}{2}\left(a_i\left(v - \sum\limits_{j=1}^{m}B_{ij}a_j\right)\right)\frac{\partial^2}{\partial{y_i}^2}P(\mathbf{y},t)\right)$}

@@

\frac{d}{dt}\mathbb{E}[y_i] = w_i(\mathbf{a})\mathbb{E}[y_i] + a_i\sum\limits_{k=1}^{m}B_{ik}\mathbb{E}[y_k]

@@

	(r + \sum\limits_{j=1}^{m}B_{1j}a_j + a_1B_{11}) & a_1B_{12} & a_1B_{13} & \dots & \dots & \dots & a_1B_{1m}\\
	a_2B_{21} & (r + \sum\limits_{j=1}^{m}B_{2j}a_j + a_2B_{22}) & a_2B_{23} & \dots & \dots & \dots & a_2B_{2m}\\

@@

	a_{i}B_{i1} & a_iB_{i2} & a_iB_{i3} & \dots & (r + \sum\limits_{j=1}^{m}B_{ij}a_j + a_iB_{ii}) & \dots & a_iB_{im}\\

@@

	a_mB_{m1} & a_mB_{m2} & a_mB_{m3} & \dots & \dots & \dots & (r + \sum\limits_{j=1}^{m}B_{mj}a_j + a_mB_{mm})

@@

The eigenvalues of the first matrix on the RHS will tell us whether the fixed point $\mathbb{E}[y_i] = 0 \ \forall \ i$ (the only fixed point of this system) is stable, or whether fluctuations are expected to grow (up to the point where the fluctuations are so large that the WNA is no longer valid).

@@

\section{An example for quantitative traits: The quantitative logistic equation}

Consider the birth and death functionals given by \eqref{Rogers_logistic_BD}. In terms of the scaled variable $\phi = K\nu$, these functions read:
\begin{equation}
\label{Rogers_logistic_BD_scaled}
\begin{aligned}
b_K(x|\phi) &= \frac{1}{K}b(x|\nu) = \frac{1}{K}\left( r\int\limits_{\mathcal{T}}m(x,y)K\phi(y)dy\right)\\
    d_K(x|\phi) &= \frac{1}{K}d(x|\nu) =  \frac{1}{K}\left(\frac{K\phi(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)K\phi(y)dy\right)
\end{aligned}
\end{equation}
Thus, using equation \eqref{deterministic_traj}, the deterministic trajectory becomes:
\begin{equation}
\label{Rogers_logistic_BD_deterministic}
\frac{\partial \psi}{\partial t}(x,t) = r\int\limits_{\mathcal{T}}m(x,y)\psi(y,t)dy-\frac{1}{n(x)}\psi(x,t)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y,t)dy
\end{equation}
Note that if we employ the change of variables $\Psi = K\psi$ to go back from $\mathcal{M}_{K}$ (\textit{i.e} $\phi^{(t)}$) to $\mathcal{M}$ (\textit{i.e} $\nu^{(t)}$), we recover the familiar continuous logistic equation (Equation \eqref{cts_logistic}) as the deterministic limit:
\begin{align*}
\frac{\partial \Psi}{\partial t}(x,t) &= r\int\limits_{\mathcal{T}}m(x,y)\Psi(y,t)dy-\frac{\Psi(x,t)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\Psi(y,t)dy \\
&\approx r\Psi(x,t) -\frac{\Psi(x,t)}{K(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\Psi(y,t)dy + D_m\nabla^2_{x}\Psi(x,t)
\end{align*}
where $K(x) = Kn(x)$ is the carrying capacity experienced by an individual of phenotype $x$, and $D_m = r \sigma_m^2/2$ measures the `diffusion rate' of the population in trait space. It is left as an exercise for the reader to verify by the same steps that if we instead have the birth rate functional $b(x|\phi) = \lambda\int m(x,y)\phi(y)dy$ (with $m(x,y)$ as defined in \eqref{Rogers_logistic_BD}) and the death rate functional $d(x|\phi) = \phi(x)\left(\mu+(\lambda-\mu)\phi(x)/K\right)$, the macroscopic limit yields the famous Fisher-KPP equation with growth rate $r=\lambda-\mu$ and diffusion constant $D = \lambda \sigma_m^2/2$.\\
\\
In any case, for the system defined by \eqref{Rogers_logistic_BD_scaled}, we can also calculate $\mathcal{D}_{\zeta}[\mathcal{A}^-]$ as
\begin{align*}
\mathcal{D}_{\zeta}[\mathcal{A}^-] &= \frac{d}{d\epsilon}\left( r\int\limits_{\mathcal{T}} m(x,y)(\psi(y)+\epsilon\zeta(y))dy - \frac{\psi(x)+\epsilon\zeta(x)}{n(x)}\int\limits_{\mathcal{T}}\alpha(x,y)(\psi(y)+\epsilon\zeta(y))dy\right) \biggl{|}_{\epsilon = 0}\\
&= r\int\limits_{\mathcal{T}}m(x,y)\zeta(y)dy - \frac{1}{n(x)}\left(\psi(x)\int\limits_{\mathcal{T}}\alpha(x,y)\zeta(y)dy + \zeta(x)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y)dy\right)
\end{align*}

>>>>

\section{An example in one dimension: The stochastic logistic equation}
Here, we analyze example \ref{ex_1D_stoch_logistic}. To recap, we had a population of individuals that exhibit a constant per-capita birth rate $\lambda > 0$, and a per-capita death rate that had the linear density-dependence $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Thus, we have the equations
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dB_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we
obtain the logistic equation. Letting $\alpha(t)$ be the solution of the logistic equation \eqref{ex_1D_stoch_logistic_det_limit}, We can Taylor expand $A^{\pm}(x)$ for the weak noise approximation, and we find:
\begin{align*}
A^-_1(x) &= \frac{d}{dx}(rx(1-x))\biggl{|}_{x=\alpha} = r(1 - 2\alpha(t))\\
A^+_0(x) &= \alpha(t)(v+r\alpha(t))
\end{align*}
Thus, the weak noise approximation of \ref{ex_1D_stoch_logistic_BD_eqns} is given by
\begin{equation}
    X_t = \alpha(t) + \frac{1}{\sqrt{K}}Y_t
\end{equation}
where the stochastic process $Y_t$ is an Ornstein-Uhlenbeck process given by the solution to the linear SDE
\begin{align}
    dY_t &= A^-_1(t)Y_tdt + \sqrt{A^+_0(t)}dB_t\nonumber\\
    \Rightarrow dY_t &= r(1 - 2\alpha(t))Y_tdt + \sqrt{\alpha(t)(v+r\alpha(t))}dB_t\label{ex_1D_stoch_logistic_WNA}
\end{align}

@@

Let us imagine a population with $m$ types of individuals that are interacting according to some ecological rules. Let the state of the population be characterized by the vector $\mathbf{v}(t) = [v_1(t),v_2(t),\ldots,v_m(t)]$, where $v_i(t)$ is the number of type $i$ individuals at time $t$. Let the birth and death rates of the $i$th type be given by:

@@

b_i(\mathbf{v}) &= \left(\lambda + \frac{1}{K}\left(\sum\limits_{j=1}^{m}\beta_{ij}v_j\right)\right)v_i\\
d_i(\mathbf{v}) &= \left(\mu + \frac{1}{K}\left(\sum\limits_{j=1}^{m}\delta_{ij}v_j\right)\right)v_i

@@

where $K > 0$ is our system size parameter (and represents a global carrying capacity across all types), $\lambda > 0$ and $\mu > 0$ are suitable positive constants representing the baseline natality and mortality common to all types, and $\beta_{ij}$ and $\delta_{ij}$ are constants describing the effect of type $j$ individuals on the birth and death rate of type $i$ individuals respectively. The sign of $M_{ij} \coloneqq \beta_{ij} - \delta_{ij}$ determines whether type $j$ has a net positive or negative effect on the growth of type $i$. In ecological communities, this is a per-capita ecological interaction effect. In game-theoretic terms, we can interpret $M_{ij}$ as the payoff obtained by a type $j$ individual playing against a type $i$ individual. I assume that $| M_{ij} | \ll K$. The values $M_{ij}$ are often collected in an $m \times m$ matrix $\mathbf{M}$ called the `payoff matrix' (in evolutionary game theory) or `interaction matrix' (in Lotka-Volterra models). Lotka-Volterra models also frequently assume that the diagonal elements $M_{ii}$ are all equal, though I will not make that assumption here.\\

@@

b^{(K)}_i(\mathbf{x}) &= \left(\lambda + \sum\limits_{j=1}^{m}\beta_{ij}x_j\right)x_i\\
d^{(K)}_i(\mathbf{x}) &= \left(\mu + \sum\limits_{j=1}^{m}\delta_{ij}x_j\right)x_i

@@

A^{\pm}_{i} = x_i\left(\left(\lambda \pm \mu\right) + \sum\limits_{j=1}^{m}\left(\beta_{ij} \pm \delta_{ij}\right)x_j\right)

@@

Defining $r \coloneqq \lambda - \mu$, $\nu \coloneqq \lambda + \mu$, and $T_{ij} \coloneqq \beta_{ij} + \delta_{ij}$, we see from equation \eqref{nD_Ito_SDE} that the mesoscopic view is the $m$ dimensional SDE given by

@@

\mathbf{A^-}_i = {(\mathbf{X}_{t})}_i(r + \sum\limits_{j=1}^{m}M_{ij}{(\mathbf{X}_{t})}_j) 

@@

(\mathbf{D}\mathbf{D}^{\mathrm{T}})_i = {(\mathbf{X}_{t})}_i(\nu + \sum\limits_{j=1}^{m}T_{ij}{(\mathbf{X}_{t})}_j) 

@@

\frac{d x_i}{dt} = x_i\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)

@@

These are precisely the Lotka-Volterra equations for a system of $m$ species. By matching the terms of \eqref{nD_example_density_b_d_rates} with those of \eqref{nD_functional_forms_for_replicator}, we can identify that we have $\mu = 0$ and

@@

w_i(\mathbf{x}) = r + \sum\limits_{j=1}^{m}M_{ij}x_j

@@

&= \sum\limits_{i=1}^{m}\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)p_i\\
&= r + \sum\limits_{i=1}^{m}p_i\left(\sum\limits_{j=1}^{m}M_{ij}x_j\right)

@@

\frac{1}{N_K(t)}\frac{dp_i}{dt} = \left[(\mathbf{Mp})_i - \mathbf{p}\cdot\mathbf{Mp}\right]p_i   

@@

which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $M$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks. Both the stochastic dynamics \eqref{nD_example_SDE} and the deterministic limit \eqref{nD_example_det_limit} can be simplified from an $m$ dimensional system to an $m-1$ dimensional system by a coordinate transformation that projects the dynamics onto an appropriate curve: If we go from the variables $x_1,\ldots,x_m$ to the variables $p_1,\ldots,p_{m-1},N_K$, we can exploit the fact that $N_K$ varies much less than the $p_i$ terms to project the system onto a `slow manifold' in which $N_K$ is approximately constant, thus obtaining an $m-1$ dimensional system of equations and recovering the relation between the Lotka-Volterra equations for $m$ species and the replicator equation for $m-1$ tactics \citep{constable_mapping_2017,parsons_dimension_2017}. However, I will not explore such dimensional reduction techniques further in this manuscript, and refer the reader to \citep{constable_stochastic_2013} and \citep{parsons_dimension_2017} for a review of the ideas of (stochastic) dynamics on slow manifolds.\\

@@

&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_k\left(\frac{\partial}{\partial x_k}(r+\sum\limits_{j=1}^{m}M_{ij}x_j)\bigg{|}_{\mathbf{x}=\mathbf{a}(t)}\right)\\
&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_kM_{ik}\\

@@

where we have used the fact that $w_i(\mathbf{y}) = r + \sum\limits_{k=1}^{m}y_kM_{ik}$ (from \eqref{nD_example_fitness}) in the last step. Thus, in the weak noise approximation of our process, the dynamics are given by

@@

\resizebox{1.1\textwidth}{!}{$\displaystyle\frac{\partial P}{\partial t}(\mathbf{y},t) = \sum\limits_{i=1}^{m}\left(-\frac{\partial}{\partial y_i}\left\{\left(y_iw_i(\mathbf{a}) + a_iw_i(\mathbf{y}) - ra_i\right)P(\mathbf{y},t)\right\}+\frac{1}{2}\left(a_i\left(\nu + \sum\limits_{j=1}^{m}T_{ij}a_j\right)\right)\frac{\partial^2}{\partial{y_i}^2}P(\mathbf{y},t)\right)$}

@@

\frac{d}{dt}\mathbb{E}[y_i] = w_i(\mathbf{a})\mathbb{E}[y_i] + a_i\sum\limits_{k=1}^{m}M_{ik}\mathbb{E}[y_k]

@@

	(r + \sum\limits_{j=1}^{m}M_{1j}a_j + a_1M_{11}) & a_1M_{12} & a_1M_{13} & \dots & \dots & \dots & a_1M_{1m}\\
	a_2M_{21} & (r + \sum\limits_{j=1}^{m}M_{2j}a_j + a_2M_{22}) & a_2M_{23} & \dots & \dots & \dots & a_2M_{2m}\\

@@

	a_{i}M_{i1} & a_iM_{i2} & a_iM_{i3} & \dots & (r + \sum\limits_{j=1}^{m}M_{ij}a_j + a_iM_{ii}) & \dots & a_iM_{im}\\

@@

	a_mM_{m1} & a_mM_{m2} & a_mM_{m3} & \dots & \dots & \dots & (r + \sum\limits_{j=1}^{m}M_{mj}a_j + a_mM_{mm})

@@

The eigenvalues of the first matrix on the RHS will tell us whether the fixed point $\mathbb{E}[y_i] = 0 \ \forall \ i$ (the only fixed point of this system) is stable, or whether fluctuations are expected to grow (up to the point where the fluctuations are so large that the WNA is no longer valid). In the $m=2$ case, \cite{tao_stochastic_2007} have shown that $\mathbb{E}[y_i] = 0 \ \forall \ i$ is a stable fixed point for this system iff the point $\mathbf{y}$ is an ESS (in the usual game-theoretic sense) for the matrix game defined by the payoff matrix $\mathbf{M}$.

@@

\section{An example for quantitative traits: The quantitative logistic equation}
Recall the birth and death functionals given by \eqref{Rogers_logistic_BD}. That is, the functionals
\begin{equation}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
corresponding to an asexual population having a constant (per-capita) birth rate $r$ and mutations controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$). In terms of the scaled variable $\phi = K\nu$, these functions read:
\begin{equation}
\label{Rogers_logistic_BD_scaled}
\begin{aligned}
b_K(x|\phi) &= \frac{1}{K}b(x|\nu) = \frac{1}{K}\left( r\int\limits_{\mathcal{T}}m(x,y)K\phi(y)dy\right)\\
    d_K(x|\phi) &= \frac{1}{K}d(x|\nu) =  \frac{1}{K}\left(\frac{K\phi(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)K\phi(y)dy\right)
\end{aligned}
\end{equation}
Thus, using equation \eqref{deterministic_traj}, the deterministic trajectory becomes:
\begin{equation}
\label{Rogers_logistic_BD_deterministic}
\frac{\partial \psi}{\partial t}(x,t) = r\int\limits_{\mathcal{T}}m(x,y)\psi(y,t)dy-\frac{1}{n(x)}\psi(x,t)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y,t)dy
\end{equation}
Note that if we employ the change of variables $\Psi = K\psi$ to go back from $\mathcal{M}_{K}$ (\textit{i.e} $\phi^{(t)}$) to $\mathcal{M}$ (\textit{i.e} $\nu^{(t)}$), we recover the familiar continuous logistic equation (Equation \eqref{cts_logistic}) as the deterministic limit:
\begin{align*}
\frac{\partial \Psi}{\partial t}(x,t) &= r\int\limits_{\mathcal{T}}m(x,y)\Psi(y,t)dy-\frac{\Psi(x,t)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\Psi(y,t)dy \\
&\approx r\Psi(x,t) -\frac{\Psi(x,t)}{K(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\Psi(y,t)dy + D_m\nabla^2_{x}\Psi(x,t)
\end{align*}
where $K(x) = Kn(x)$ is the carrying capacity experienced by an individual of phenotype $x$, and $D_m = r \sigma_m^2/2$ measures the `diffusion rate' of the population in trait space. It is left as an exercise for the reader to verify by the same steps that if we instead have the birth rate functional $b(x|\phi) = \lambda\int m(x,y)\phi(y)dy$ (with $m(x,y)$ as defined in \eqref{Rogers_logistic_BD}) and the death rate functional $d(x|\phi) = \phi(x)\left(\mu+(\lambda-\mu)\phi(x)/K\right)$, the macroscopic limit yields the famous Fisher-KPP equation with growth rate $r=\lambda-\mu$ and diffusion constant $D = \lambda \sigma_m^2/2$.\\
\\
In any case, for the system defined by \eqref{Rogers_logistic_BD_scaled}, we can also calculate $\mathcal{D}_{\zeta}[\mathcal{A}^-]$ as
\begin{align*}
\mathcal{D}_{\zeta}[\mathcal{A}^-] &= \frac{d}{d\epsilon}\left( r\int\limits_{\mathcal{T}} m(x,y)(\psi(y)+\epsilon\zeta(y))dy - \frac{\psi(x)+\epsilon\zeta(x)}{n(x)}\int\limits_{\mathcal{T}}\alpha(x,y)(\psi(y)+\epsilon\zeta(y))dy\right) \biggl{|}_{\epsilon = 0}\\
&= r\int\limits_{\mathcal{T}}m(x,y)\zeta(y)dy - \frac{1}{n(x)}\left(\psi(x)\int\limits_{\mathcal{T}}\alpha(x,y)\zeta(y)dy + \zeta(x)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y)dy\right)
\end{align*}
